Synthesising Recursive Functions for First-Order Model Counting

### 2023-02-14 doc/paper/paper.tex
Fix references

% TODO: use \textt for predicates (e.g., p, q), introduce the notation, use DeclareMathOperator
% TODO: cite the linear order axiom paper \cite{DBLP:journals/corr/abs-2211-01164} in the introduction
% TODO: introduction: move the huge footnote to the main text

@@

% TODO: cite some works about synthesis

@@

Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the
domain(s)~\cite{jaeger2012liftability}. Over more than a decade, many classes of
instances were shown to be
liftable~\cite{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} showed that the class
of all sentences of FOL with up to two variables (denoted \FOtwo{}) is liftable.
Then Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15} proved that there exists
a sentence with three variables for which FOMC is $\#\P_{1}$-complete (i.e.,
\FOthree{} is not liftable). Since these two seminal results, most of the
research on (W)FOMC focused on developing faster solutions for the \FOtwo{}
fragment~\cite{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and
\SRU{}~\cite{DBLP:conf/nips/KazemiKBP16},
\Uone{}~\cite{DBLP:conf/lics/KuusistoL18}, \FOtwo{} with tree
axioms~\cite{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the two-variable
fragment with counting
quantifiers)~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On
the empirical front, there are several implementations of exact WFOMC
algorithms: \textsc{Alchemy}~\cite{DBLP:journals/cacm/GogateD16},
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21},
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}, and
\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16}. Approximate counting is supported by
\textsc{Alchemy}, \textsc{ApproxWFOMC}~\cite{DBLP:conf/ijcai/BremenK20},
\textsc{ForcLift}~\cite{DBLP:conf/uai/BroeckCD12},
\textsc{Magician}~\cite{DBLP:conf/aaai/VenugopalSG15}, and
\textsc{Tuffy}~\cite{DBLP:journals/pvldb/NiuRDS11}.

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{There is some previous work
  showing how FOMC can compute
  factorials~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/kr/BremenK21,DBLP:conf/ijcai/Broeck16}.
  One of them~\cite{DBLP:conf/ijcai/Broeck16} is an example `by hand' that does
  not directly correspond to any algorithm.
  Another~\cite{DBLP:journals/jair/Kuzelka21} proves that the problem is
  liftable but contains no experimental results. The
  third~\cite{DBLP:conf/kr/BremenK21} contains experimental results performed
  with an algorithm, which is not publicly available.} We claim that this
shortcoming is due to the inability of these algorithms to construct recursive
solutions. The topic of recursion in the context of WFOMC has been studied
before but in limited ways. Barv{\'{\i}}nek et
al.~\shortcite{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to generate numerical
data that is then used to conjecture recurrence relations that explain that
data. Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} introduced the idea of
\emph{domain recursion}. Intuitively, domain recursion partitions a domain of
size $n$ into a single explicitly named constant and the remaining domain of
size $n-1$. However, many stringent conditions are enforced to ensure that the
search for a tractable solution always terminates.

@@

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift}. \textsc{ForcLift} works in two
stages: compilation and evaluation/propagation. In the first stage, various
\emph{(compilation) rules} are applied to the input (or some derivative)
formula, gradually constructing a circuit. In the second stage, the weights of
the instance are propagated through the circuit, computing the weighted model
count. Along with new compilation rules, \textsc{Crane} introduces changes to
both stages of the process. First, while \textsc{ForcLift} applies compilation
rules via greedy search, \textsc{Crane} also supports a hybrid search algorithm
that applies some rules greedily and some using breadth-first search. Second,
the product of compilation is not directly evaluated but rather interpreted as a
collection of functions on domain sizes.

@@

\begin{example}\label{example:smokers}
  Consider a variant of the well-known `friends and smokers' example
  $\forall X, Y \in \Delta\text{.
  } \smokes(X) \land \friends(X, Y) \Rightarrow \smokes(Y)$. Letting
  $n \coloneqq |\Delta|$ as in \cref{example:simple}, the model count can be
  expressed as
  $\sum_{k=0}^{n} \binom{n}{k}2^{n^{2} - k(n-k)}$~\cite{DBLP:conf/kr/BroeckMD14}.
\end{example}
% TODO: maybe add an extra sentence to explain what the formula means

>>>>

@@

@@

Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the
domain(s)~\cite{jaeger2012liftability}. Over more than a decade, many classes of
instances were shown to be
liftable~\cite{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} showed that the class
of all sentences of FOL with up to two variables (denoted \FOtwo{}) is liftable.
Then Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15} proved that there exists
a sentence with three variables for which FOMC is $\#\P_{1}$-complete (i.e.,
\FOthree{} is not liftable). Since these two seminal results, most of the
research on (W)FOMC focused on developing faster solutions for the \FOtwo{}
fragment~\cite{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and
\SRU{}~\cite{DBLP:conf/nips/KazemiKBP16},
\Uone{}~\cite{DBLP:conf/lics/KuusistoL18}, \Ctwo{} (i.e., the two-variable
fragment with counting
quantifiers)~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}, and
\Ctwo{} extended with tree axioms~\cite{DBLP:conf/kr/BremenK21} and linear order
axiom~\cite{DBLP:journals/corr/abs-2211-01164}. On the empirical front, there
are several implementations of exact WFOMC algorithms:
\textsc{Alchemy}~\cite{DBLP:journals/cacm/GogateD16},
\textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21},
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}, and
\textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16}. Approximate counting is supported by
\textsc{Alchemy}, \textsc{ApproxWFOMC}~\cite{DBLP:conf/ijcai/BremenK20},
\textsc{ForcLift}~\cite{DBLP:conf/uai/BroeckCD12},
\textsc{Magician}~\cite{DBLP:conf/aaai/VenugopalSG15}, and
\textsc{Tuffy}~\cite{DBLP:journals/pvldb/NiuRDS11}.

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{There is some previous work
  showing how FOMC can compute
  factorials~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/kr/BremenK21,DBLP:conf/ijcai/Broeck16}.
  One of them~\cite{DBLP:conf/ijcai/Broeck16} is an example `by hand' that does
  not directly correspond to any algorithm.
  Another~\cite{DBLP:journals/jair/Kuzelka21} proves that the problem is
  liftable but contains no experimental results. The
  third~\cite{DBLP:conf/kr/BremenK21} contains experimental results performed
  with an algorithm, which is not publicly available.} We claim that the
capabilities of (W)FOMC algorithms can be significantly extended by empowering
them with the ability to construct recursive solutions. The topic of recursion
in the context of WFOMC has been studied before but in limited ways.
Barv{\'{\i}}nek et al.~\shortcite{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to
generate numerical data that is then used to conjecture recurrence relations
that explain that data. Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11}
introduced the idea of \emph{domain recursion}. Intuitively, domain recursion
partitions a domain of size $n$ into a single explicitly named constant and the
remaining domain of size $n-1$. However, many stringent conditions are enforced
to ensure that the search for a tractable solution always terminates.

@@

In this work, we show how to relax these restrictions in a way that results in a
stronger (W)FOMC algorithm, capable of handling more instances (e.g., counting
various injective mappings) in a lifted manner. The ideas presented in this
paper are implemented in \textsc{Crane}---an extension of the arguably most
well-known WFOMC algorithm \textsc{ForcLift}. \textsc{ForcLift} works in two
stages: compilation and evaluation/propagation. In the first stage, various
\emph{(compilation) rules} are applied to the input (or some derivative)
formula, gradually constructing a circuit. In the second stage, the weights of
the instance are propagated through the circuit, computing the weighted model
count. Along with new compilation rules, \textsc{Crane} introduces changes to
both stages of the process. First, while \textsc{ForcLift} applies compilation
rules via greedy search, \textsc{Crane} also supports a hybrid search algorithm
that applies some rules greedily and some using breadth-first search. Second,
the product of compilation is not directly evaluated but rather interpreted as a
collection of functions on domain sizes. Hence, our approach is reminiscent of
previous work on lifted inference via compilation to C++
programs~\cite{DBLP:conf/kr/KazemiP16} and the broader area of functional
synthesis~\cite{DBLP:conf/cav/GoliaRM20,DBLP:conf/pldi/KuncakMPS10,sanathanan1963transfer}.

% similarly to the other domains, there are many functions that fit the
% specification, i.e., can correctly count the number of models as a function of
% domain size(s). Whereas in [...] one typically wants to select the smallest
% definition of a function, we are primarily concerned with its worst-case
% complexity as measured by the number of arithmetic operations needed for an
% arbitrary function call
% \begin{itemize}
%   \item Boolean functions~\cite{DBLP:conf/cav/GoliaRM20}
%   \item dynamical systems~\cite{sanathanan1963transfer}
%   \item program fragments~\cite{DBLP:conf/pldi/KuncakMPS10}
% \end{itemize}

@@

\begin{example}\label{example:smokers}
  Consider a variant of the well-known `friends and smokers' example
  $\forall X, Y \in \Delta\text{.
  } \smokes(X) \land \friends(X, Y) \Rightarrow \smokes(Y)$. Letting
  $n \coloneqq |\Delta|$ as in \cref{example:simple}, the model count can be
  expressed as
  $\sum_{k=0}^{n} \binom{n}{k}2^{n^{2} - k(n-k)}$~\cite{DBLP:conf/kr/BroeckMD14}.
\end{example}
